### 2019-05-20 Writing/response_to_reviewers.tex

\section{Editors' comments}

\begin{enumerate}
\label{metric_change}

\item{\textcolor{blue}{Along the lines set out by Reviewer 2, it would be wise to ensure that your metric is characterised as thoroughly as possible. In this regard, does it meet the following five conditions?}}

After working on the required proof of the metric, we figured out that the formerly proposed Character Difference had a non-expected behaviour of being non-linear.
We have now modified the metric to be linear (see illustration here: \url{https://github.com/TGuillerme/CharactersCorrelation/blob/master/Analysis/05-char.diffModification.pdf}).
We thus made some major changes to the manuscript that were not suggested by the reviewers below:
\begin{enumerate}
    \item We change the Character Difference metric to be a scaled Hamming Distance rather than the former proposed distance.
    \item We reran all the simulations from scratch with the inclusion of the updated Character Difference metric.
    \item For time constraint reasons we reran 20 replicates (rather than 35 in the previous manuscript) for a total of 40 CPU years.
    \item We propose a new interpretation to the updated results.
\end{enumerate}

@@

\begin{enumerate}

\item{\textcolor{blue}{I have some concerns about the approach, however. I was somewhat confused by the CD metric. When quantifying n, how is it decided how many taxa have "comparable" characters?
If you simulated the character for all taxa, do they not all have this character (i.e. n always equals the dataset size)?
But more importantly, will two characters not be more similar if they share the same number of character states, independent of if they are telling you the same evolutionary history?
It seems to me that whether or not two characters share a state space is not the relevant axis along which to measure correlation.}}
\label{gower}

Using a Gower based distance, we only compared the comparable taxa between characters.
For example, for two characters \texttt{x = {0, 1, 2, 3}} and \texttt{y = {0, ?, ?, ?}}, the Character Difference metric will be effectively calculated using only the first taxon (\texttt{x = {0}} and \texttt{y = {0}}) and would result in a Character Difference of 0.
As the reviewer points out, characters with less taxa in common will thus be less likely to be different.
This is consistent with what the CD metric actually captures in terms of phylogenetic signal (the number of share splits).
In the example above, the both characters imply at least one split (between the first taxon and the others).
In other words, in a phylogenetic software, both characters will support the same split.

As the reviewer also correctly points out, we simulated our data with no missing data so this does not affect our results.
The precision in the manuscript was just describing the properties of the new CD metric.

@@

\begin{enumerate}

\item{\textcolor{blue}{"Phylogenetic analysis algorithms require the assumption of character independence - a condition generally acknowledged to be violated by morphological data."\\
I don't love this sentence because it emphasizes morphology as distinct, when we recognize non-independence as a problem in basically all data types (see Huelsenbeck and Nielsen, 1996, Effect of Nonindependent Substitution on Phylogenetic Accuracy and the references therein). I do think you should point to a citation here; Davalos et al. 2014 (Integrating Incomplete Fossils by Isolating Conflicting Signal in Saturated and Non-Independent Morphological Characters) is a good one.}}
\label{molecular}

We slightly rephrased the sentence in the abstract (See reviewer 3 minor revision \ref{abstract}) and expanded on the distinction between both character types this in the introduction following the reviewer pertinent biobliography suggestion:

\textit{This independence is assumed for both molecular and morphological characters (Huelsenbeck and Nielsen, 1999; D\'{a}valos et al., 2014, Zou and Zhang, 2016).
However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differs in size and in the amount of states analysed (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but sometimes up to more than ten; Guillerme and Cooper, 2016).
Second, this difference between both types of characters has been suggested to be stronger correlations in morphological data (i.e. ``concentrated'' correlations) \textit{versus} weaker and more ``spread'' correlations in molecular data (i.e. ``diffused'' correlations - though some regions of the genome can be thought as more correlated than others; Huelsenbeck and Nielsen, 1999).
Furthermore, because of this ``spread'' correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function cf. changes in nucleotide states ; see below).} lines 43-57.

@@

\item{\textcolor{blue}{"Evolutionary dependence: this is the result of sets of characters co-evolving due to selection, likely related to functional links between two traits that help serve an overall lifestyle trait."\\
Seems appropriate to cite Clarke and Middleton (2008) here.}}

We added the suggested citation (lines 75).

\item{\textcolor{blue}{Coding dependence:\\
This section could use a figure illustrating these types of dependence, and what they mean. For example: "For instance, coding for the same absence in different characters creates state transformations associated with the loss or gain of a particular character." is a fairly confusing sentence. I took this to mean that if you are coding, for example, tail characters, you might atomize the tail into a suite of characters, all of which are absent in organisms with no tail. But I'm not sure I've understood your meaning correctly.}}

We do not feel that adding a figure for illustrating the matrix will improve clarity of an already long and thorough description.
However, we've trying to clarify the description with an example as follows:

\textit{For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail.} lines 95-96 and \textit{In the example above, attributes of the tail (colour or length) are hierarchically dependent on the presence of the tail in the first place.} lines 102-104.

@@

\item{\textcolor{blue}{"Correlation between characters detected by the software:"\\
I really struggled with this section. It seems like the problem here is still a coding problem - that the authors of these four characters have chosen four highly dependent characters to build the matrix. Even scoring by hand under parsimony would have problems here. I think what you need to decide here is what you want to communicate in this paragraph. Is it that software is not currently equipped to handle this issue?}}

We've added a paragraph highlighting how the problem indeed comes from the fact that current software are not equipped to handle this issue:

\textit{In fact, for practicality reasons, phylogenetic software do not (or rarely) distinguish between the different sources of correlation.
In this specific example, we would intuitively want to favour character \textbf{C1} and \textbf{C2} that seem to be more likely to reflect the split by separating artiodactyles from the cetaceans.
In other words, this is the correlation between characters that depicts the evolutionary history of this group.
However, the other characters depict the same history (i.e. the same split) but for the wrong reasons (note that even if \textbf{C2} was coded presence (1) or absence (0) of baleen, the same split would still be estimated by the software).
One solution would be to a) weight some characters (e.g. giving \textbf{C3} and \textbf{C4} a lower weight) and b) make some characters dependent on each other (e.g. making \textbf{C3} and \textbf{C4} dependent on \textbf{C1}).
However, this can also be a bad practice since it can easily lead to the resulting tree reflecting the researcher's prior expectations rather than the true evolutionary history.} lines 143-154.

@@

\item{\textcolor{blue}{"To assess the effects of character correlation on the accuracy of phylogenetic inference we generated a series of matrices exhibiting different levels of correlation between some characters (Fig.1 - note that each step is described in more details below):"\\
Fig. 1 is the results, but this text implies it is a workflow diagram.}}

We double checked the figure links and made sure that this mention of Fig.1 now properly points to the workflow figure.

@@

\item{\textcolor{blue}{Most of the time when RF is normalized, it's done so that low scores are better - i.e. a low RF score indicates less distance to the true tree. I was very confused looking at the figures, until I realized this was not how normalizing had been performed. }}

We've changed the labels of the figures to NTS(RF) and NTS(Tr) to avoid any confusions.

\item{\textcolor{blue}{"However, the effect of these sources of correlation was out of the scope of this study and would have required a posteriori changes to the matrices which are - when using empirical data - at best bad practice and at worth dishonest."\\
I'm not sure I follow this.}}
\label{dishonest}

We meant that modifying phylogenetic matrices after the phylogenetic inference is a way to test the effect of the sources of correlation but is probably not a good practice in systematics.
Although re-weighting characters based on consistency or retention indices can be used to objectively improve the topology, changing the matrix's characters \textit{a posteriori}, it can be seen as dishonest to remove characters because they ``don't give the right signal''). We've removed the end of this sentence however to avoid any further confusion.

@@

\item{\textcolor{blue}{"However, we do not compare the ``maximised'', ``minimised'' and ``randomised'' to the ``true'' tree but rather to the ``normal'' tree. "\\
I understand why you did this, but I think you should also compare normal to true. That will give us some idea of the scope of how close we can get to true, under these conditions, and will help contextualize the results.}}

We've added the comparisons between the ``normal'' and ``randomised'' trees to the ``true'' tree in the supplementary material 3, figures 9 to 11.
We've added a reference to these figures in the main text:

\textit{However, we did report the comparison between the ``true'' trees and both the ``normal'' and ``randomised'' in the supplementary materials 3, Figs 12, 13 and 14.} lines 344-346

@@

\subsection{Major suggestions}

\begin{enumerate}

\item{\textcolor{blue}{1. To be a metric, there are several properties that need to be satisfied. The proposed function
is positively valued and symmetric (copied below from line 178): [...] But, the following metric property does not hold for all x; y:}}
\label{proof}

We are really grateful that this reviewer picked up this flaw in our mathematical demonstration of the CD metric.
We have reviewed the formula of the CD metric following an unexpected non-linear behaviour (see comment \ref{metric_change}).
Furtheremore, we'd like to highlight the following point:

In fact, before being plugged into the CD equation, characters need to be modified following Felsenstein's \textit{xyz} notation.
In other words, because the character state tokens are non-metric values if the characters are unordered they can be translated in a standard way so that the first token in the character is renamed to ``1'', the second to ``2'' etc.
For example, in a 3 state character for four taxa: \texttt{X = {1, 0, 2, 1}}, the character tokens that are integers representing the states could equally have been \texttt{X = {E, A, I, E}} or \texttt{X = {blue, red, absent, blue}}.
In order to numerically compare X to any other character it is thus more simple to translate both characters in a standard way such as the first token is replaced by the symbol ``1'', etc.

For example if comparing two characters \texttt{X = {1, 0, 2, 1}} and \texttt{Y = {2, 1, 0, 2}}, the absolute difference would be \texttt{X - Y = {-1, -1, 2, -1}}, or when using a Hamming distance \texttt{X - Y = {1, 1, 1, 1}} showing maximum difference between both characters.
However, because these tokens are non-numeric and effectively represent the same phylogenetic split, both characters are translated to \texttt{X' = {1, 2, 3, 1}} and \texttt{Y' = {1, 2, 3, 1}} leading to a difference of \texttt{X' - Y' = {0, 0, 0,0}}.

In the reviewer's demonstration, \texttt{x  = {1}} and \texttt{y  = {0}} thus become \texttt{x' = {1}} and \texttt{y' = {1}}.
Hence $CD_{(x',y')} = 0 \Leftrightarrow x' = y'$.
This is also reflected in the new $CD$ formula that makes comparing two characters for a single taxon nonsensical: the denominator is equal to $0$ if $n = 1$.

Furthermore, the CD metric, used in this manuscript as a proxy for character correlation actually measures the difference between in characters in terms of phylogenetic signal.
The character \texttt{X = {1, 0, 2, 1, 0}} described above implies two splits (and of course same goes for the translated \texttt{X' = {1, 2, 3, 1, 2}}).
Another character \texttt{Y = {1, 1, 1, 2, 2}} will imply only one split, not shared with \texttt{X} and a character \texttt{Z = {0, 2, 1, 0, 2}} (\texttt{Z' = {1, 2, 3, 1, 2}}) would imply the same split as \texttt{X} (hence, $CD_{(X',Z')} = 0$ and $CD_{(X',Y')} = 0.75$).
In other words the probability of X and Z generating conflicting split signal is null and this probability for X and Y is of 0.75.

### 2019-05-20 Writing/draft_highlighted.tex
Updated Martin's comments

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data \cite[e.g.][]{Davalos01072014, ZouConvergence}.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
\hl{This independence is assumed for both molecular and morphological characters} \citep{huelsenbeck1999effect,Davalos01072014,ZouConvergence}.
\hl{However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differs in size and in the amount of states analysed (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but sometimes up to more than ten;}
\citealt{Guillerme2016146}
\hl{).
Second, this discrepancy has been suggested to be stronger correlations in morphological data (i.e. ``concentrated'' correlations) \textit{versus} weaker and more ``spread'' correlations in molecular data}
\citep[\hl{i.e. ``diffused'' correlations - though some regions of the genome can be thought as more correlated than others}][]{huelsenbeck1999effect}.
\hl{Furthermore, because of this ``spread'' correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function \textit{cf.} changes in nucleotide states ; see below).}
Before discussing character correlation further, it is important to understand that it may manifest itself in at least three distinct ways:

@@

    \item \textbf{Coding dependence:} this is the results of researcher methodology for defining or/and coding discrete morphological characters \citep{Brazeau2011,simoes2017giant}.
    Coding dependence manifests itself in several ways, particularly in coding redundant information.
    For instance, coding for the same absence in different characters creates state transformations associated with the loss or gain of a particular character.
    This occurs when a number of multistate characters include two variable feature states (e.g. large, small; red, blue etc.) in conjunction with absence.
    \hl{For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail.}
    It is worth noting, however, that these correlations could also be due to the nature of the available data, especially in palaeontology.
    For example, when only one fragmentary molar is available to describe a specimen, researchers have to ``extract'' as much phylogenetic information from the available data as possible, potentially inducing correlations.
    This coding dependency is linked to hierarchical dependency between characters \citep{wilkinson1995coping,BrazeauNA}.
    Finally this can also be due to a bias in the amount of characters available.
    For example, in skulls, because of their complexity, there is a high likelihood of inducing correlation (by effectively reducing structural complexity to discrete characters).
\end{itemize}

@@

In the example in Table \ref{Tab:example_matrix}, the characters \textbf{C1} and \textbf{C2} are the most likely to be truly independent; characters \textbf{C3} and \textbf{C4} suffer from a a coding induced dependency; characters \textbf{C1} and \textbf{C3}/\textbf{C4} have an evolutionary induce dependency and again, characters \textbf{C2} and \textbf{C3}/\textbf{C4} are likely to be independent.
Yet a phylogenetic software will treat all these four characters in exactly the same way: only the sheer difference between the character states tokens will be used in order to infer the tree.
\hl{In fact, for practicality reasons, phylogenetic software do not (or rarely) distinguish between the different sources of correlation.
In this specific example, we would intuitively want to favour character \textbf{C1} and \textbf{C2} more than the other characters since this would be \textit{a priori} the characters that are the most likely to reflect the group's evolutionary history in terms of split by separating the artiodactyles from the cetaceans.
In other words, this is the correlation between characters that depicts the evolutionary history of this group.
However, the other characters depict the same history (i.e. the same split) but for the wrong reasons (note that even if \textbf{C2} was coded presence (1) or absence (0) of baleens, the same split would still be estimated by the software).
One solution could be to weight characters \textit{a priori} but that is a bad practice since it can easily lead to the resulting tree reflecting the researcher's \textit{priori} rather than the true evolutionary history.}
% \hl{Furthermore, phenotypic variations can be continuous rather than discrete and ``discretising'' such characters can have some influence on tree inference} \citep{parins2017use}.
% \hl{In this case, ``discretised'' }.
This illustrates how some characters will therefore be expected to covary in non-phylogenetic way, and that this phenomenon can reasonably be expected to mislead phylogenetic analysis.
Yet the question has never been explored through a thorough simulation framework (although it has been tackled empirically for morphological data \citealt{Davalos01072014} or molecular data \citealt{ZouConvergence}).

@@

\subsection{Comparing topologies}
We compared the topologies using the same approach as in \cite{Guillerme2016146}: we measured both the Robinson-Foulds distance \citep{RF1981} and the triplets distance \citep{dobson1975triplets} between the trees inferred from the ``maximised'', ``minimised'' and ``randomised'' matrices and the tree inferred from the ``normal'' matrix.
We explored the effect of character difference on recovering the ``normal'' topology by comparing the ``maximised'', ``minimised'' and ``randomised'' trees to the ``normal'' tree (Figs \ref{Fig:RF_results_best} and \ref{Fig:Tr_results_best} and supplementary materials 3 Figs 7 and 8).
Note that we are not comparing the trees to the ``true'' tree used to simulate the matrices.
First, in biology, this tree is always unknown.
Second, our objective is to measure the direct effect of character correlation approximated by the difference in topology between the ``normal'', ``maximised'' and ``minimised'' trees.
When measuring the difference between these trees and the ``true'' tree, we would also confound the effect of simulating a birth-death tree and simulating a discrete morphological matrices from it.
\hl{However, we did report the comparison between the ``true'' trees and ``normal'' and ``randomised'' ones is available in the supplementary materials 3, Figs 12, 13 and 14.}

>>>>

The non-independence of large numbers of morphological characters is often cited in anticipation of problems with morphological data \cite[e.g.][]{Davalos01072014, ZouConvergence}.
The assumption of character independence is central to phylogenetic inference methods such as maximum likelihood and maximum parsimony \citep[e.g.][]{joysey1982problems,felsenstein1985phylogenies,lewisa2001,felsenstein2004inferring}.
\hl{This independence is assumed for both molecular and morphological characters} \citep{huelsenbeck1999effect,Davalos01072014,ZouConvergence}.
\hl{However, several distinctions can be made between the amount and the nature of the correlation.
First, molecular and morphological data often differs in size and in the amount of states analysed (molecular matrices are generally larger with mainly four states - not including missing data or indels - and morphological matrices are smaller with generally two states but sometimes up to more than ten;}
\citealt{Guillerme2016146}
\hl{).
Second, this difference between both types of characters has been suggested to be stronger correlations in morphological data (i.e. ``concentrated'' correlations) \textit{versus} weaker and more ``spread'' correlations in molecular data}
\citep[\hl{i.e. ``diffused'' correlations - though some regions of the genome can be thought as more correlated than others}][]{huelsenbeck1999effect}.
\hl{Furthermore, because of this ``spread'' correlations in molecular data, it can be harder to link the character correlation to a specific biological phenomenon (i.e. changes in morphological states can be more readily linked to a function \textit{cf.} changes in nucleotide states ; see below).}
Before discussing character correlation further, it is important to understand that it may manifest itself in at least three distinct ways:

@@

    \item \textbf{Coding dependence:} this is the results of researcher methodology for defining or/and coding discrete morphological characters \citep{Brazeau2011,simoes2017giant}.
    Coding dependence manifests itself in several ways, particularly in coding redundant information.
    For instance, coding for the same absence in different characters creates state transformations associated with the loss or gain of a particular character.
    This occurs when a number of multistate characters include two variable feature states (e.g. large, small; red, blue etc.) in conjunction with absence.
    \hl{For example, two characters ``tail colour'' and ``tail length'' could be coded two times as an absence for a taxon with no tail.}
    It is worth noting, however, that these correlations could also be due to the nature of the available data, especially in palaeontology.
    For example, when only one fragmentary molar is available to describe a specimen, researchers have to ``extract'' as much phylogenetic information from the available data as possible, potentially inducing correlations.
    This coding dependency is linked to hierarchical dependency between characters \citep{wilkinson1995coping,BrazeauNA}.
    \hl{In the example above, attributes of the tail (colour or length) are hierarchically dependent on the presence of the tail in the first place.}
    Finally this can also be due to a bias in the amount of characters available.
    For example, in skulls, because of their complexity, there is a high likelihood of inducing correlation (by effectively reducing structural complexity to discrete characters).
\end{itemize}

@@

In the example in Table \ref{Tab:example_matrix}, the characters \textbf{C1} and \textbf{C2} are the most likely to be truly independent; characters \textbf{C3} and \textbf{C4} suffer from a a coding induced dependency; characters \textbf{C1} and \textbf{C3}/\textbf{C4} have an evolutionary induce dependency and again, characters \textbf{C2} and \textbf{C3}/\textbf{C4} are likely to be independent.
Yet a phylogenetic software will treat all these four characters in exactly the same way: only the sheer difference between the character states tokens will be used in order to infer the tree.
\hl{In fact, for practicality reasons, phylogenetic software do not (or rarely) distinguish between the different sources of correlation.
In this specific example, we would intuitively want to favour character \textbf{C1} and \textbf{C2} that seem to be more likely to reflect the split by separating artiodactyles from the cetaceans.
In other words, this is the correlation between characters that depicts the evolutionary history of this group.
However, the other characters depict the same history (i.e. the same split) but for the wrong reasons (note that even if \textbf{C2} was coded presence (1) or absence (0) of baleen, the same split would still be estimated by the software).
One solution would be to a) weight some characters (e.g. giving \textbf{C3} and \textbf{C4} a lower weight) and b) make some characters dependent on each other (e.g. making \textbf{C3} and \textbf{C4} dependent on \textbf{C1}).
However, this can also be a bad practice since it can easily lead to the resulting tree reflecting the researcher's prior expectations rather than the true evolutionary history.}
This illustrates how some characters will therefore be expected to covary in non-phylogenetic way, and that this phenomenon can reasonably be expected to mislead phylogenetic analysis.
Yet the question has never been explored through a thorough simulation framework (although it has been tackled empirically for morphological data \citealt{Davalos01072014} or molecular data \citealt{ZouConvergence}).

@@

\subsection{Comparing topologies}
We compared the topologies using the same approach as in \cite{Guillerme2016146}: we measured both the Robinson-Foulds distance \citep{RF1981} and the triplets distance \citep{dobson1975triplets} between the trees inferred from the ``maximised'', ``minimised'' and ``randomised'' matrices and the tree inferred from the ``normal'' matrix.
We explored the effect of character difference on recovering the ``normal'' topology by comparing the ``maximised'', ``minimised'' and ``randomised'' trees to the ``normal'' tree (Figs \ref{Fig:RF_results_best} and \ref{Fig:Tr_results_best} and supplementary materials 3 Figs 7 and 8).
Note that we are not comparing the trees to the ``true'' tree used to simulate the matrices.
First, in biology, this tree is always unknown.
Second, our objective is to measure the direct effect of character correlation approximated by the difference in topology between the ``normal'', ``maximised'' and ``minimised'' trees.
When measuring the difference between these trees and the ``true'' tree, we would also confound the effect of simulating a birth-death tree and simulating a discrete morphological matrices from it.
\hl{However, we did report the comparison between the ``true'' trees and both the ``normal'' and ``randomised'' in the supplementary materials 3, Figs 12, 13 and 14.}
